---
title: "Clinical Pathway Discovery with HealthProcessAI"
author: "SMAILE Lab, Karolinska Institutet"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Clinical Pathway Discovery with HealthProcessAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
```

# Introduction

This vignette demonstrates how to use HealthProcessAI's R implementation for discovering and analyzing clinical pathways. We'll explore how to:

- Load and prepare clinical event data
- Discover process models using bupaR
- Identify pathway variants and bottlenecks
- Analyze patient flow patterns
- Generate actionable clinical insights

## Background

Clinical pathway discovery uses process mining techniques to analyze healthcare event logs and uncover:

- **Standard care pathways** - Common sequences of clinical activities
- **Pathway variants** - Alternative routes through care processes
- **Process bottlenecks** - Points where delays frequently occur
- **Resource utilization** - How clinical resources are used across pathways
- **Quality indicators** - Metrics for pathway performance

# Setup and Data Loading

First, let's load the required packages and initialize our HealthProcessAI components:

```{r libraries, message=FALSE, warning=FALSE}
# Load required libraries
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lubridate)
  library(knitr)
  library(DT)
})

# Source HealthProcessAI R modules
source("../core/step1_data_loader.R")
source("../core/step2_process_mining.R") 
source("../core/step4_advanced_analytics.R")
source("../core/report_generator.R")
```

## Create Sample Emergency Department Data

For this demonstration, we'll create a realistic dataset representing emergency department patient flows:

```{r create_sample_data}
create_ed_pathway_data <- function(n_patients = 200) {
  set.seed(42)  # For reproducibility
  patients <- data.frame()
  
  for (i in 1:n_patients) {
    case_id <- paste0("ED", sprintf("%04d", i))
    start_time <- as.POSIXct("2024-01-01") + days(sample(1:90, 1)) + 
                  hours(sample(0:23, 1)) + minutes(sample(0:59, 1))
    
    # Patient characteristics
    age <- sample(18:95, 1)
    severity <- sample(c("Minor", "Moderate", "Severe", "Critical"), 1,
                      prob = c(0.4, 0.3, 0.2, 0.1))
    
    # Generate pathway based on severity
    pathway <- generate_pathway(severity, age)
    
    # Create events for this patient
    current_time <- start_time
    for (j in 1:nrow(pathway)) {
      activity <- pathway$activity[j]
      resource <- pathway$resource[j]
      duration_mins <- pathway$duration_mins[j]
      
      patients <- rbind(patients, data.frame(
        case = case_id,
        activity = activity,
        timestamp = current_time,
        resource = resource,
        age = age,
        severity = severity,
        duration_mins = duration_mins,
        stringsAsFactors = FALSE
      ))
      
      current_time <- current_time + minutes(duration_mins)
    }
  }
  
  return(patients)
}

generate_pathway <- function(severity, age) {
  base_pathway <- data.frame(
    activity = c("Registration", "Triage", "Initial_Assessment"),
    resource = c("Registration", "Nurse", "Physician"),
    duration_mins = c(5, 10, 20),
    stringsAsFactors = FALSE
  )
  
  # Add activities based on severity
  if (severity == "Minor") {
    additional <- data.frame(
      activity = c("Treatment", "Discharge"),
      resource = c("Nurse", "Physician"),
      duration_mins = c(30, 10),
      stringsAsFactors = FALSE
    )
  } else if (severity == "Moderate") {
    additional <- data.frame(
      activity = c("Lab_Tests", "Radiology", "Treatment", "Observation", "Discharge"),
      resource = c("Lab", "Radiology", "Physician", "Nurse", "Physician"),
      duration_mins = c(45, 30, 60, 120, 15),
      stringsAsFactors = FALSE
    )
  } else if (severity == "Severe") {
    additional <- data.frame(
      activity = c("Lab_Tests", "Radiology", "Specialist_Consult", 
                  "Treatment", "Extended_Observation", "Ward_Admission"),
      resource = c("Lab", "Radiology", "Specialist", "Physician", "Nurse", "Admin"),
      duration_mins = c(30, 25, 45, 90, 240, 30),
      stringsAsFactors = FALSE
    )
  } else { # Critical
    additional <- data.frame(
      activity = c("Immediate_Treatment", "Lab_Tests", "Radiology", 
                  "ICU_Consult", "ICU_Admission"),
      resource = c("Emergency_Team", "Lab", "Radiology", "ICU_Physician", "Admin"),
      duration_mins = c(15, 20, 15, 30, 45),
      stringsAsFactors = FALSE
    )
  }
  
  return(rbind(base_pathway, additional))
}

# Generate the sample dataset
ed_data <- create_ed_pathway_data(n_patients = 200)

# Display basic information about the dataset
cat("Emergency Department Dataset Overview:\n")
cat("- Total events:", nrow(ed_data), "\n")
cat("- Unique patients:", length(unique(ed_data$case)), "\n")
cat("- Date range:", as.character(range(ed_data$timestamp)), "\n")
cat("- Activities:", length(unique(ed_data$activity)), "\n")
cat("- Resources:", length(unique(ed_data$resource)), "\n")

# Show first few rows
head(ed_data) %>% kable(caption = "Sample Emergency Department Event Data")
```

# Pathway Discovery and Analysis

## Step 1: Load and Validate Data

```{r load_data}
# Initialize the data loader
loader <- EventLogLoader$new()

# Convert our sample data to event log format
event_log <- loader$convert_to_event_log(ed_data)

# Display validation results
validation_results <- loader$validate_event_log(event_log)
cat("Data Validation Results:\n")
cat("- Valid cases:", validation_results$valid_cases, "\n")
cat("- Valid events:", validation_results$valid_events, "\n")
cat("- Data quality score:", round(validation_results$quality_score, 3), "\n")
```

## Step 2: Process Mining and Pathway Discovery

```{r process_mining}
# Initialize the process miner
miner <- ProcessMiner$new()

# Discover the Directly-Follows Graph (DFG)
dfg_result <- miner$discover_dfg(event_log)

cat("Process Discovery Results:\n")
cat("- Unique activities:", length(dfg_result$activities), "\n")
cat("- Process transitions:", nrow(dfg_result$transitions), "\n")
cat("- Start activities:", length(dfg_result$start_activities), "\n")
cat("- End activities:", length(dfg_result$end_activities), "\n")

# Display most frequent transitions
top_transitions <- dfg_result$transitions %>%
  arrange(desc(frequency)) %>%
  head(10)

kable(top_transitions, caption = "Top 10 Most Frequent Process Transitions")
```

## Step 3: Pathway Variant Analysis

```{r pathway_variants}
# Identify process variants
variants <- miner$identify_variants(event_log)

cat("Pathway Variant Analysis:\n")
cat("- Total variants found:", length(variants$variants), "\n")
cat("- Most common variant covers", 
    round(max(variants$variant_frequencies) * 100, 1), "% of cases\n")

# Display top variants
top_variants <- variants$variant_summary %>%
  head(8)

kable(top_variants, caption = "Most Common Clinical Pathway Variants")

# Visualize variant distribution
variants$variant_summary %>%
  head(10) %>%
  ggplot(aes(x = reorder(variant_id, frequency), y = frequency)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Emergency Department Pathway Variants",
    subtitle = "Distribution of most common patient journey patterns",
    x = "Pathway Variant",
    y = "Number of Patients",
    caption = "Data: Simulated ED patient flows"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray60"),
    axis.text.y = element_text(size = 8)
  )
```

## Step 4: Performance Analysis

```{r performance_analysis}
# Calculate performance metrics
performance <- miner$calculate_performance_metrics(event_log)

cat("Process Performance Metrics:\n")
cat("- Average case duration:", round(performance$avg_case_duration, 1), "minutes\n")
cat("- Median case duration:", round(performance$median_case_duration, 1), "minutes\n")
cat("- Case duration std dev:", round(performance$std_case_duration, 1), "minutes\n")

# Activity-level performance
activity_performance <- performance$activity_metrics
kable(activity_performance, caption = "Performance Metrics by Activity")

# Visualize case duration distribution by severity
ed_data %>%
  group_by(case, severity) %>%
  summarise(
    total_duration = sum(duration_mins, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = severity, y = total_duration, fill = severity)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Emergency Department Length of Stay by Severity",
    subtitle = "Distribution of total patient journey duration",
    x = "Patient Severity Level",
    y = "Total Duration (minutes)",
    fill = "Severity"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(type = "qual", palette = "Set2")
```

## Step 5: Bottleneck Identification

```{r bottleneck_analysis}
# Identify process bottlenecks
bottlenecks <- miner$identify_bottlenecks(event_log)

cat("Bottleneck Analysis Results:\n")
cat("- Activities analyzed:", length(bottlenecks$activity_bottlenecks), "\n")
cat("- Critical bottlenecks identified:", 
    sum(bottlenecks$activity_bottlenecks$is_bottleneck), "\n")

# Display bottleneck activities
bottleneck_activities <- bottlenecks$activity_bottlenecks %>%
  filter(is_bottleneck == TRUE) %>%
  arrange(desc(avg_waiting_time))

kable(bottleneck_activities, caption = "Identified Process Bottlenecks")

# Visualize bottlenecks
bottlenecks$activity_bottlenecks %>%
  arrange(desc(avg_waiting_time)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(activity, avg_waiting_time), y = avg_waiting_time)) +
  geom_col(aes(fill = is_bottleneck), alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Activity Performance Analysis",
    subtitle = "Average waiting time by clinical activity",
    x = "Clinical Activity",
    y = "Average Waiting Time (minutes)",
    fill = "Is Bottleneck"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("FALSE" = "lightblue", "TRUE" = "coral"))
```

# Advanced Analytics

## Resource Utilization Analysis

```{r resource_analysis}
# Analyze resource utilization
resource_analysis <- ed_data %>%
  group_by(resource, severity) %>%
  summarise(
    patient_count = n_distinct(case),
    total_duration = sum(duration_mins, na.rm = TRUE),
    avg_duration = mean(duration_mins, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_duration))

kable(head(resource_analysis, 12), caption = "Resource Utilization by Severity Level")

# Visualize resource workload
resource_analysis %>%
  group_by(resource) %>%
  summarise(
    total_patients = sum(patient_count),
    total_time = sum(total_duration),
    .groups = "drop"
  ) %>%
  arrange(desc(total_time)) %>%
  head(8) %>%
  ggplot(aes(x = reorder(resource, total_time), y = total_time)) +
  geom_col(fill = "darkblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Resource Utilization in Emergency Department",
    subtitle = "Total time allocated by resource type",
    x = "Resource Type",
    y = "Total Time (minutes)",
    caption = "Higher values indicate heavier resource utilization"
  ) +
  theme_minimal()
```

## Temporal Pattern Analysis

```{r temporal_patterns}
# Analyze temporal patterns
temporal_patterns <- ed_data %>%
  mutate(
    hour_of_day = hour(timestamp),
    day_of_week = wday(timestamp, label = TRUE),
    is_weekend = wday(timestamp) %in% c(1, 7)
  ) %>%
  group_by(hour_of_day, severity) %>%
  summarise(
    patient_count = n_distinct(case),
    .groups = "drop"
  )

# Visualize hourly patient patterns
temporal_patterns %>%
  ggplot(aes(x = hour_of_day, y = patient_count, fill = severity)) +
  geom_area(alpha = 0.7, position = "stack") +
  labs(
    title = "Emergency Department Patient Arrivals by Hour",
    subtitle = "Daily pattern of patient presentations by severity",
    x = "Hour of Day",
    y = "Number of Patients",
    fill = "Severity Level"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  scale_fill_brewer(type = "qual", palette = "Set1")
```

# Clinical Insights and Recommendations

## Pathway Efficiency Metrics

```{r efficiency_metrics}
# Calculate pathway efficiency metrics
efficiency_metrics <- ed_data %>%
  group_by(case, severity) %>%
  summarise(
    pathway_length = n(),
    total_duration = sum(duration_mins),
    unique_resources = n_distinct(resource),
    .groups = "drop"
  ) %>%
  group_by(severity) %>%
  summarise(
    avg_pathway_length = round(mean(pathway_length), 1),
    avg_duration = round(mean(total_duration), 1),
    avg_resources = round(mean(unique_resources), 1),
    efficiency_score = round(avg_pathway_length / (avg_duration / 60), 2),
    .groups = "drop"
  )

kable(efficiency_metrics, caption = "Pathway Efficiency Metrics by Severity Level")
```

## Quality Indicators

```{r quality_indicators}
# Calculate quality indicators
quality_indicators <- list(
  total_patients = length(unique(ed_data$case)),
  avg_length_of_stay = round(mean(aggregate(duration_mins ~ case, ed_data, sum)$duration_mins), 1),
  pathway_compliance = round(length(variants$variants) / length(unique(ed_data$case)) * 100, 1),
  resource_efficiency = round(nrow(ed_data) / length(unique(ed_data$resource)), 1)
)

cat("Quality Indicators Summary:\n")
for (indicator in names(quality_indicators)) {
  cat("-", indicator, ":", quality_indicators[[indicator]], "\n")
}
```

# Actionable Recommendations

Based on our pathway discovery analysis, here are key recommendations:

## 1. **Bottleneck Resolution**
```{r recommendations_bottlenecks, echo=FALSE}
if (nrow(bottleneck_activities) > 0) {
  cat("Priority bottlenecks to address:\n")
  for (i in 1:min(3, nrow(bottleneck_activities))) {
    activity <- bottleneck_activities$activity[i]
    wait_time <- round(bottleneck_activities$avg_waiting_time[i], 1)
    cat(paste0("- ", activity, " (avg wait: ", wait_time, " minutes)\n"))
  }
} else {
  cat("No critical bottlenecks identified in current analysis.\n")
}
```

## 2. **Resource Optimization**
- **Peak hour staffing**: Increase resources during high-volume hours (typically 10 AM - 6 PM)
- **Severity-based allocation**: Allocate specialist resources based on patient acuity patterns
- **Cross-training**: Train staff on multiple activities to improve flexibility

## 3. **Process Improvements**
- **Pathway standardization**: Develop standard protocols for the top 5 pathway variants
- **Parallel processing**: Enable concurrent activities where clinically appropriate
- **Early assessment**: Implement rapid assessment protocols for high-severity cases

## 4. **Quality Metrics**
- **Target length of stay**: Aim for severity-specific duration targets
- **Pathway compliance**: Monitor adherence to established clinical protocols
- **Resource utilization**: Optimize resource allocation based on demand patterns

# Conclusion

This analysis demonstrates how HealthProcessAI's R implementation can effectively:

1. **Discover clinical pathways** from event log data
2. **Identify process variants** and their frequencies  
3. **Detect bottlenecks** and inefficiencies
4. **Analyze resource utilization** patterns
5. **Generate actionable insights** for process improvement

## Next Steps

1. **Real data integration**: Replace sample data with actual clinical event logs
2. **Conformance checking**: Compare actual pathways against established protocols
3. **Predictive modeling**: Build models to predict pathway outcomes
4. **Continuous monitoring**: Implement ongoing pathway performance monitoring
5. **Clinical validation**: Validate findings with clinical domain experts

---

**About SMAILE Lab**

This vignette was developed at SMAILE (Stockholm Medical Artificial Intelligence and Learning Environments), Karolinska Institutet. For more information, visit [https://smile.ki.se](https://smile.ki.se).

**HealthProcessAI Repository**: [https://github.com/ki-smile/HealthProcessAI](https://github.com/ki-smile/HealthProcessAI)