---
title: "Getting Started with HealthProcessAI in R"
author: "SMAILE Lab, Karolinska Institutet"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with HealthProcessAI in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150
)
```

# Welcome to HealthProcessAI R Implementation

HealthProcessAI provides a comprehensive R framework for healthcare process mining with integrated Large Language Model (LLM) capabilities. This getting started guide will walk you through:

- Installation and setup
- Basic concepts and terminology
- Your first process mining analysis
- Integration with LLMs for clinical insights
- Next steps and advanced features

## What is Healthcare Process Mining?

Process mining is a data science technique that uses event logs to:

- **Discover** actual care pathways from healthcare data
- **Analyze** process performance and identify bottlenecks
- **Monitor** compliance with clinical protocols
- **Improve** healthcare processes through data-driven insights

HealthProcessAI combines traditional process mining with AI-powered analysis to generate actionable clinical insights.

# Installation and Setup

## 1. Install Required Packages

```{r install_packages, eval=FALSE}
# Install required CRAN packages
install.packages(c(
  "tidyverse", "R6", "httr2", "jsonlite", 
  "lubridate", "cluster", "rmarkdown"
))

# Install suggested packages for enhanced functionality
install.packages(c(
  "bupaR", "edeaR", "processmapR", "ggplot2", 
  "testthat", "styler", "lintr"
))
```

## 2. Load HealthProcessAI

```{r load_packages, message=FALSE, warning=FALSE}
# Load required libraries
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(ggplot2)
  library(knitr)
})

# Source HealthProcessAI R modules
# Note: In a real installation, these would be loaded via library(HealthProcessAI)
source("../core/step1_data_loader.R")
source("../core/step2_process_mining.R") 
source("../core/step3_llm_integration.R")
source("../core/step4_advanced_analytics.R")
source("../core/step5_orchestrator.R")
source("../core/report_generator.R")

cat("✅ HealthProcessAI R modules loaded successfully!\n")
```

## 3. Verify Installation

```{r verify_installation}
# Verify core classes can be instantiated
tryCatch({
  loader <- EventLogLoader$new()
  miner <- ProcessMiner$new()
  generator <- ReportGenerator$new()
  
  cat("✅ All core components initialized successfully!\n")
  cat("🚀 Ready to start process mining!\n")
}, error = function(e) {
  cat("❌ Error during initialization:", conditionMessage(e), "\n")
})
```

# Basic Concepts and Terminology

## Key Components

HealthProcessAI follows a **5-step modular pipeline**:

1. **Data Loading** (`EventLogLoader`) - Load and validate healthcare event logs
2. **Process Mining** (`ProcessMiner`) - Discover clinical pathways using bupaR
3. **LLM Integration** (`LLMAnalyzer`) - Generate AI-powered insights
4. **Advanced Analytics** (`AdvancedProcessAnalyzer`) - Clustering and predictions
5. **Report Orchestration** (`ReportOrchestrator`) - Multi-model consensus building

## Event Log Format

Healthcare event logs contain:
- **Case**: Unique patient/episode identifier
- **Activity**: Clinical activity or event name  
- **Timestamp**: When the activity occurred
- **Resource**: Who performed the activity (optional)
- **Additional attributes**: Patient demographics, severity, etc.

# Your First Process Mining Analysis

Let's create and analyze a simple healthcare dataset to demonstrate the core workflow.

## Step 1: Create Sample Data

```{r create_sample_data}
# Create sample emergency department data
create_sample_ed_data <- function(n_patients = 50) {
  set.seed(123)  # For reproducibility
  events <- data.frame()
  
  for (i in 1:n_patients) {
    case_id <- sprintf("ED_%03d", i)
    arrival_time <- as.POSIXct("2024-01-01") + 
                    days(sample(1:30, 1)) + 
                    hours(sample(8:20, 1))
    
    # Define pathway based on patient severity
    severity <- sample(c("Low", "Medium", "High"), 1, prob = c(0.5, 0.3, 0.2))
    
    if (severity == "Low") {
      pathway <- data.frame(
        activity = c("Arrival", "Triage", "Consultation", "Discharge"),
        duration_mins = c(0, 10, 30, 10)
      )
    } else if (severity == "Medium") {
      pathway <- data.frame(
        activity = c("Arrival", "Triage", "Consultation", "Tests", "Treatment", "Discharge"),
        duration_mins = c(0, 10, 20, 45, 60, 15)
      )
    } else {  # High severity
      pathway <- data.frame(
        activity = c("Arrival", "Triage", "Emergency_Treatment", 
                    "Tests", "Specialist_Consult", "Admission"),
        duration_mins = c(0, 5, 15, 30, 45, 30)
      )
    }
    
    # Generate events with timestamps
    current_time <- arrival_time
    for (j in 1:nrow(pathway)) {
      events <- rbind(events, data.frame(
        case = case_id,
        activity = pathway$activity[j],
        timestamp = current_time,
        resource = paste0("Resource_", sample(1:5, 1)),
        severity = severity,
        stringsAsFactors = FALSE
      ))
      current_time <- current_time + minutes(pathway$duration_mins[j])
    }
  }
  
  return(events)
}

# Generate sample data
sample_data <- create_sample_ed_data(n_patients = 100)

cat("Sample Emergency Department Data:\n")
cat("- Total events:", nrow(sample_data), "\n")
cat("- Unique patients:", length(unique(sample_data$case)), "\n")
cat("- Activities:", length(unique(sample_data$activity)), "\n")
cat("- Date range:", paste(range(as.Date(sample_data$timestamp)), collapse = " to "), "\n")

# Display first few rows
head(sample_data) %>% kable(caption = "Sample Emergency Department Events")
```

## Step 2: Load and Validate Data

```{r load_validate_data}
# Initialize data loader
loader <- EventLogLoader$new()

# Convert to standard event log format
event_log <- loader$convert_to_event_log(sample_data)

# Validate the data
validation <- loader$validate_event_log(event_log)

cat("Data Validation Results:\n")
cat("- Quality Score:", round(validation$quality_score, 2), "/100\n")
cat("- Valid Cases:", validation$valid_cases, "\n")
cat("- Valid Events:", validation$valid_events, "\n")

if (length(validation$issues) > 0) {
  cat("- Issues Found:", length(validation$issues), "\n")
  for (issue in validation$issues[1:min(3, length(validation$issues))]) {
    cat("  •", issue, "\n")
  }
}
```

## Step 3: Discover Process Patterns

```{r discover_process}
# Initialize process miner
miner <- ProcessMiner$new()

# Discover the Directly-Follows Graph
dfg_result <- miner$discover_dfg(event_log)

cat("Process Discovery Results:\n")
cat("- Unique Activities:", length(dfg_result$activities), "\n")
cat("- Process Transitions:", nrow(dfg_result$transitions), "\n")

# Show top transitions
cat("\nMost Frequent Process Transitions:\n")
top_transitions <- dfg_result$transitions %>%
  arrange(desc(frequency)) %>%
  head(8)

kable(top_transitions, caption = "Most Common Activity Transitions")
```

## Step 4: Analyze Process Variants

```{r analyze_variants}
# Identify process variants (different patient pathways)
variants <- miner$identify_variants(event_log)

cat("Pathway Variants Analysis:\n")
cat("- Total Variants:", length(variants$variants), "\n")
cat("- Most Common Variant:", 
    round(max(variants$variant_frequencies) * 100, 1), "% of patients\n")

# Display variant summary
variant_summary <- variants$variant_summary %>%
  head(6)

kable(variant_summary, caption = "Most Common Patient Pathways")

# Visualize variant distribution
variants$variant_summary %>%
  head(8) %>%
  ggplot(aes(x = reorder(variant_id, frequency), y = frequency)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Emergency Department Patient Pathway Variants",
    subtitle = "Distribution of different patient journey patterns",
    x = "Pathway Variant",
    y = "Number of Patients"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "gray60")
  )
```

## Step 5: Performance Analysis

```{r performance_analysis}
# Calculate performance metrics
performance <- miner$calculate_performance_metrics(event_log)

cat("Process Performance Overview:\n")
cat("- Average Case Duration:", round(performance$avg_case_duration, 1), "minutes\n")
cat("- Median Case Duration:", round(performance$median_case_duration, 1), "minutes\n")

# Activity-level performance
activity_perf <- performance$activity_metrics %>%
  arrange(desc(avg_duration))

kable(head(activity_perf), caption = "Activity Performance Metrics")

# Visualize case durations by severity
sample_data %>%
  group_by(case, severity) %>%
  summarise(
    start_time = min(timestamp),
    end_time = max(timestamp),
    total_duration = as.numeric(difftime(end_time, start_time, units = "mins")),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = severity, y = total_duration, fill = severity)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Emergency Department Length of Stay by Severity",
    x = "Patient Severity",
    y = "Total Duration (minutes)",
    fill = "Severity"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# LLM Integration for Clinical Insights

*Note: LLM features require an OpenRouter API key. Get yours at [https://openrouter.ai](https://openrouter.ai)*

## Setting up LLM Integration

```{r llm_setup, eval=FALSE}
# Set your API key (do this once per session)
Sys.setenv(OPENROUTER_API_KEY = "your-api-key-here")

# Initialize LLM analyzer
analyzer <- LLMAnalyzer$new(api_key = Sys.getenv("OPENROUTER_API_KEY"))

# Check available models
models <- analyzer$get_available_models()
head(models[c("id", "name", "pricing")]) %>% kable()
```

## Generate Clinical Insights

```{r llm_analysis, eval=FALSE}
# Create clinical analysis prompt
process_summary <- list(
  total_patients = length(unique(event_log$case)),
  avg_duration = round(performance$avg_case_duration, 1),
  top_activities = head(dfg_result$activities, 5),
  main_pathways = head(variants$variant_summary$pathway, 3)
)

prompt <- analyzer$create_clinical_prompt(
  process_summary = process_summary,
  context_info = list(
    setting = "Emergency Department",
    timeframe = "1 month analysis",
    focus = "process optimization"
  )
)

# Query Claude for insights
insights <- analyzer$query_model(
  model_name = "anthropic/claude-3.5-sonnet",
  prompt = prompt,
  temperature = 0.3
)

cat("Clinical Insights Generated:\n")
cat("- Response length:", nchar(insights$response), "characters\n")
cat("- Tokens used:", insights$usage$total_tokens, "\n")
cat("- Model:", insights$model, "\n")

# Display insights (first 500 characters)
cat("\nInsights Preview:\n")
cat(substr(insights$response, 1, 500), "...\n")
```

# Advanced Features

## Generate Comprehensive Report

```{r generate_report}
# Initialize report generator
generator <- ReportGenerator$new()

# Prepare results for reporting
analysis_results <- list(
  event_log = event_log,
  dfg_result = dfg_result,
  variants = variants,
  performance = performance,
  metadata = list(
    analysis_date = Sys.Date(),
    dataset_info = "Emergency Department Sample Data",
    analyst = "HealthProcessAI Tutorial"
  )
)

# Generate markdown report
report_content <- generator$generate_markdown_report(
  results = analysis_results,
  title = "Emergency Department Process Analysis",
  include_visualizations = TRUE
)

# Save report (optional)
# writeLines(report_content, "ed_analysis_report.md")

cat("✅ Comprehensive report generated!\n")
cat("Report sections:", length(strsplit(report_content, "##")[[1]]), "\n")
cat("Report length:", nchar(report_content), "characters\n")
```

## Health Check

```{r health_check, eval=FALSE}
# Run comprehensive health check
source("../comprehensive_health_check.R")

checker <- ComprehensiveHealthChecker$new()
health_report <- checker$run_all_checks()

cat("Health Check Summary:\n")
cat("- Total Checks:", health_report$summary$total_checks, "\n")
cat("- Passed:", health_report$summary$passed, "\n")
cat("- Failed:", health_report$summary$failed, "\n")
cat("- Warnings:", health_report$summary$warnings, "\n")
cat("- Overall Status:", health_report$summary$status, "\n")
```

# Next Steps

Congratulations! You've completed your first HealthProcessAI analysis. Here are recommended next steps:

## 1. **Explore Real Data**
- Load your own healthcare event logs
- Use `EventLogLoader$load_from_csv()` for CSV files
- Try `EventLogLoader$load_from_excel()` for Excel data

## 2. **Advanced Analytics**
```{r next_steps_advanced, eval=FALSE}
# Initialize advanced analyzer
advanced <- AdvancedProcessAnalyzer$new(event_log)

# Cluster patient pathways
clusters <- advanced$cluster_patient_pathways(n_clusters = 3)

# Calculate clinical KPIs
kpis <- advanced$calculate_clinical_kpis()

# Predictive monitoring
predictions <- advanced$predictive_monitoring()
```

## 3. **Multi-Model LLM Analysis**
```{r next_steps_llm, eval=FALSE}
# Compare insights from multiple models
comparison <- analyzer$multi_model_analysis(
  models = c("anthropic/claude-3.5-sonnet", "google/gemini-pro"),
  prompt = clinical_prompt
)

# Use report orchestrator for consensus
orchestrator <- ReportOrchestrator$new()
consolidated <- orchestrator$consolidate_reports(comparison$responses)
```

## 4. **Integration with bupaR**
```{r next_steps_bupar, eval=FALSE}
# Convert to bupaR for advanced process mining
bupar_log <- loader$convert_to_bupar(event_log)

# Use bupaR directly for specialized analysis
library(bupaR)
library(processmapR)

# Create process map
bupar_log %>% process_map()

# Analyze resource utilization
bupar_log %>% resource_frequency()
```

## 5. **Custom Visualizations**
```{r next_steps_viz, eval=FALSE}
# Create custom process visualizations
process_map <- miner$visualize_process_map(dfg_result)
print(process_map)

# Timeline analysis
timeline_plot <- advanced$create_timeline_analysis(event_log)
print(timeline_plot)
```

# Getting Help

## Documentation and Resources

- **Package Help**: Use `?EventLogLoader`, `?ProcessMiner`, etc.
- **Vignettes**: Browse `browseVignettes("HealthProcessAI")` for more tutorials
- **Examples**: Check the `R/examples/` directory for complete workflows
- **Tests**: Review `R/tests/` for usage examples and edge cases

## Community and Support

- **GitHub Repository**: [https://github.com/ki-smile/HealthProcessAI](https://github.com/ki-smile/HealthProcessAI)
- **SMAILE Lab**: [https://smile.ki.se](https://smile.ki.se)
- **Issues**: Report bugs and request features on GitHub

## Additional Vignettes

- `clinical-pathway-discovery.Rmd` - Advanced pathway analysis
- `sepsis-analysis-tutorial.Rmd` - Sepsis-specific workflows
- `llm-integration-guide.Rmd` - Comprehensive LLM usage
- `performance-optimization.Rmd` - Optimizing analysis performance

---

**Happy Process Mining!** 🏥📊🚀

*This tutorial was developed at SMAILE (Stockholm Medical Artificial Intelligence and Learning Environments), Karolinska Institutet.*