---
title: "Getting Started with HealthProcessAI"
author: "SMAILE Lab, Karolinska Institutet"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with HealthProcessAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

HealthProcessAI is a comprehensive R framework for healthcare process mining with integrated Large Language Model (LLM) capabilities. This vignette provides a step-by-step guide to get you started with analyzing clinical pathways, patient journeys, and healthcare processes.

## Installation

```{r eval=FALSE}
# Install from GitHub (when public)
devtools::install_github("ki-smile/HealthProcessAI", subdir = "R")

# Or install locally
devtools::install_local("path/to/HealthProcessAI/R")

# Load the package
library(HealthProcessAI)
```

```{r}
# For this vignette, we'll load required packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(R6)
  library(lubridate)
})

# Source the modules (in actual package installation, this wouldn't be needed)
# source("../core/step1_data_loader.R")
# source("../core/step2_process_mining.R")
# source("../core/step3_llm_integration.R")
```

## Quick Start: 5-Step Pipeline

HealthProcessAI implements a modular 5-step pipeline for healthcare process mining:

1. **Data Loading** - Load and validate healthcare event logs
2. **Process Mining** - Discover clinical pathways
3. **LLM Integration** - Generate AI-powered insights
4. **Advanced Analytics** - Clustering and predictive analysis
5. **Report Orchestration** - Multi-model consensus building

### Step 1: Load Your Data

```{r}
# Create sample sepsis data for demonstration
sample_data <- create_sample_sepsis_data(n_patients = 20, sepsis_rate = 0.3)

# Initialize the data loader
loader <- EventLogLoader$new()

# Load the data
event_log <- loader$load_from_data_frame(sample_data)

# Get basic statistics
stats <- loader$get_basic_statistics(event_log)
print(stats)
```

### Step 2: Process Mining Analysis

```{r eval=FALSE}
# Initialize process miner
miner <- ProcessMiner$new()

# Discover process patterns (requires bupaR)
if (requireNamespace("bupaR", quietly = TRUE)) {
  # Convert to bupaR format
  bupar_log <- loader$convert_to_bupar(event_log)
  
  # Discover directly-follows graph
  dfg <- miner$discover_dfg(bupar_log, type = "frequency")
  
  # Calculate performance metrics
  metrics <- miner$calculate_performance_metrics(bupar_log)
  print(metrics)
  
  # Identify bottlenecks
  bottlenecks <- miner$identify_bottlenecks(bupar_log)
  head(bottlenecks)
}
```

### Step 3: LLM Integration (Optional)

```{r eval=FALSE}
# Initialize LLM analyzer (requires API key)
api_key <- Sys.getenv("OPENROUTER_API_KEY")

if (nchar(api_key) > 0) {
  llm_analyzer <- LLMAnalyzer$new(api_key = api_key)
  
  # Query a single model
  response <- llm_analyzer$query_model(
    model = "anthropic/claude-3.5-sonnet",
    prompt = "Analyze this sepsis progression pattern for clinical insights...",
    max_tokens = 500
  )
  
  cat(response$content)
  
  # Multi-model analysis
  multi_analysis <- llm_analyzer$analyze_with_multiple_models(
    process_results = metrics,
    models = c("anthropic/claude-3.5-sonnet", "openai/gpt-4o"),
    clinical_context = "sepsis"
  )
} else {
  message("Set OPENROUTER_API_KEY environment variable to enable LLM features")
}
```

### Step 4: Advanced Analytics

```{r}
# Initialize advanced analyzer
advanced_analyzer <- AdvancedProcessAnalyzer$new(event_log)

# Patient pathway clustering
clustering_results <- advanced_analyzer$cluster_patient_pathways(n_clusters = 3)
print(clustering_results$summary)

# Calculate clinical KPIs
kpis <- advanced_analyzer$calculate_clinical_kpis()
print(kpis)

# Predictive monitoring
predictions <- advanced_analyzer$predictive_monitoring(
  target_activity = "Sepsis_Confirmed",
  threshold = 0.3
)
head(predictions)
```

### Step 5: Report Generation

```{r eval=FALSE}
# Initialize report generator
generator <- ReportGenerator$new()

# Prepare comprehensive report data
report_data <- list(
  title = "Sepsis Patient Journey Analysis",
  date = Sys.Date(),
  author = "Clinical Analytics Team",
  executive_summary = "Analysis of sepsis progression patterns reveals key intervention opportunities.",
  
  process_analysis = list(
    n_cases = stats$n_cases,
    n_activities = stats$n_activities,
    avg_case_duration = stats$avg_case_duration,
    bottlenecks = bottlenecks
  ),
  
  clinical_insights = list(
    key_findings = c(
      "Three distinct sepsis progression pathways identified",
      "Early vital sign deterioration predicts sepsis onset"
    ),
    recommendations = c(
      "Implement automated early warning system",
      "Reduce laboratory processing delays"
    )
  ),
  
  advanced_analytics = clustering_results,
  kpis = kpis
)

# Generate markdown report
markdown_report <- generator$generate_markdown_report(report_data)

# Save to file
writeLines(markdown_report, "sepsis_analysis_report.md")

# Generate HTML report (requires rmarkdown)
if (requireNamespace("rmarkdown", quietly = TRUE)) {
  html_result <- generator$generate_html_report(
    data = report_data,
    output_file = "sepsis_analysis_report.html"
  )
}
```

## Working with Real Data

### Loading CSV Files

```{r eval=FALSE}
# Load real sepsis data
loader <- EventLogLoader$new()

# Standard format
event_log <- loader$load_from_csv("path/to/sepsis_data.csv")

# Custom column mapping
event_log <- loader$load_from_csv(
  "path/to/patient_data.csv",
  case_col = "patient_id",
  activity_col = "event_name",
  timestamp_col = "event_time", 
  resource_col = "department"
)
```

### Data Validation and Preprocessing

```{r}
# Validate event log structure
is_valid <- loader$validate_event_log(event_log)
print(paste("Data is valid:", is_valid))

# Apply filters
filtered_log <- loader$apply_filters(
  event_log,
  case_filter = c("Patient_001", "Patient_005", "Patient_010"),
  activity_filter = c("Admission", "Lab_Test_Abnormal", "Sepsis_Confirmed")
)

print(paste("Filtered to", nrow(filtered_log), "events"))
```

## Data Transformation Examples

### PhysioNet-Style Data

```{r eval=FALSE}
# Transform raw clinical measurements to event logs
transformer <- HealthcareDataTransformer$new()

# Create sample PhysioNet-style data
raw_data <- transformer$create_physionet_style_data(n_patients = 10)

# Transform to event log
transformed_log <- transformer$transform_to_event_log(raw_data)

# Analyze transformation results
analysis <- transformer$analyze_event_log(transformed_log)
print(analysis)
```

### Organ Failure Progression

```{r eval=FALSE}
# Analyze organ failure patterns
organ_transformer <- PhysioNetToOrganTransformer$new()

# Create sample patients with organ dysfunction
sample_patients <- create_sample_patients()

# Transform to organ failure events
organ_log <- organ_transformer$create_aggregated_organ_log(sample_patients)

# Analyze patterns
patterns <- organ_transformer$analyze_organ_patterns(organ_log)
print(patterns)
```

## Visualization and Analysis

### Basic Process Statistics

```{r}
# Activity frequency analysis
activity_freq <- event_log %>%
  count(activity, sort = TRUE) %>%
  head(10)

print(activity_freq)

# Case duration analysis  
case_durations <- event_log %>%
  group_by(case) %>%
  summarise(
    duration_hours = as.numeric(difftime(max(timestamp), min(timestamp), units = "hours")),
    n_activities = n(),
    .groups = "drop"
  )

summary(case_durations)
```

### Sepsis Analysis

```{r}
# Analyze sepsis vs non-sepsis pathways
sepsis_analysis <- event_log %>%
  group_by(case, sepsis_label) %>%
  summarise(
    pathway = paste(activity, collapse = " -> "),
    duration_hours = as.numeric(difftime(max(timestamp), min(timestamp), units = "hours")),
    n_activities = n(),
    .groups = "drop"
  )

# Compare sepsis vs non-sepsis cases
sepsis_comparison <- sepsis_analysis %>%
  group_by(sepsis_label) %>%
  summarise(
    n_cases = n(),
    avg_duration = mean(duration_hours, na.rm = TRUE),
    avg_activities = mean(n_activities, na.rm = TRUE),
    .groups = "drop"
  )

print(sepsis_comparison)
```

## Best Practices

### Data Preparation

1. **Validate your data** using `validate_event_log()` before analysis
2. **Handle missing values** appropriately for your clinical context
3. **Standardize activity names** for consistent analysis
4. **Check timestamp formats** and timezone consistency

### Process Mining

1. **Start simple** with basic statistics before advanced analysis
2. **Validate clinical relevance** of discovered patterns
3. **Consider clinical context** when interpreting bottlenecks
4. **Use domain expertise** to guide analysis parameters

### LLM Integration

1. **Set up API keys** securely using environment variables
2. **Start with single models** before multi-model analysis
3. **Validate AI insights** with clinical experts
4. **Monitor API costs** and usage

### Reporting

1. **Tailor reports** to your audience (clinical vs administrative)
2. **Include visualizations** to support findings
3. **Provide actionable recommendations**
4. **Document methodology** and assumptions

## Next Steps

- Explore the full API documentation for advanced features
- Try the complete pipeline examples in `R/examples/`
- Set up LLM integration with your OpenRouter API key
- Customize analysis parameters for your specific use case
- Integrate with your existing clinical data systems

## Support and Resources

- **GitHub Repository**: https://github.com/ki-smile/HealthProcessAI
- **SMAILE Lab**: https://smile.ki.se  
- **bupaR Documentation**: https://www.bupar.net/
- **OpenRouter API**: https://openrouter.ai/docs

For questions and support, please open an issue on the GitHub repository.