---
title: "Sepsis Analysis with HealthProcessAI"
author: "SMAILE Lab, Karolinska Institutet"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Sepsis Analysis with HealthProcessAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This tutorial demonstrates how to use HealthProcessAI for comprehensive sepsis analysis, from raw clinical data to AI-powered insights and actionable recommendations. We'll walk through a complete sepsis progression analysis workflow using the R implementation.

## Background

Sepsis is a life-threatening condition that requires rapid identification and treatment. Process mining can help healthcare teams understand:

- **Patient journey patterns** leading to sepsis
- **Critical intervention points** and timing
- **Resource bottlenecks** that delay care
- **Risk factors** for sepsis development
- **Quality improvement opportunities**

## Data Setup

```{r message=FALSE}
# Load required libraries
library(tidyverse)
library(lubridate)
library(R6)

# Source HealthProcessAI modules
# Note: In actual package installation, these would be loaded automatically
source("../core/step1_data_loader.R")
source("../core/step2_process_mining.R") 
source("../core/step3_llm_integration.R")
source("../core/step4_advanced_analytics.R")
source("../core/step5_orchestrator.R")
source("../core/report_generator.R")
```

### Creating Sample Sepsis Data

For this tutorial, we'll create realistic sepsis progression data:

```{r}
# Create comprehensive sepsis dataset
create_realistic_sepsis_data <- function(n_patients = 100, sepsis_rate = 0.25) {
  set.seed(123)  # For reproducibility
  
  sepsis_patients <- sample(1:n_patients, size = floor(n_patients * sepsis_rate))
  all_events <- list()
  
  # Define progression pathways
  normal_pathway <- c(
    "ED_Arrival", "Triage", "Initial_Assessment", "Vitals_Normal",
    "Lab_Orders", "Lab_Results_Normal", "Physician_Evaluation",
    "Treatment_Plan", "Discharge_Planning", "Discharge"
  )
  
  sepsis_pathway <- c(
    "ED_Arrival", "Triage", "Initial_Assessment", "Vitals_Concerning",
    "Urgent_Lab_Orders", "Lab_Results_Abnormal", "SIRS_Criteria_Met",
    "Infection_Suspected", "Blood_Cultures", "Antibiotic_Orders",
    "IV_Access", "Fluid_Resuscitation", "ICU_Consultation",
    "Sepsis_Protocol_Initiated", "Lactate_Elevated", "Vasopressor_Consideration",
    "ICU_Transfer", "Sepsis_Confirmed", "Organ_Support", "Recovery_Phase",
    "Step_Down", "Discharge"
  )
  
  for (i in 1:n_patients) {
    patient_id <- sprintf("Patient_%03d", i)
    is_sepsis <- i %in% sepsis_patients
    
    # Choose pathway and add some variation
    if (is_sepsis) {
      base_activities <- sepsis_pathway
      # Add complications for some patients
      if (runif(1) < 0.3) {
        insert_pos <- which(base_activities == "ICU_Transfer")
        base_activities <- append(base_activities, "Mechanical_Ventilation", insert_pos)
      }
      if (runif(1) < 0.2) {
        insert_pos <- which(base_activities == "Organ_Support")
        base_activities <- append(base_activities, "Dialysis_Initiated", insert_pos)
      }
    } else {
      base_activities <- normal_pathway
      # Some normal patients might have extended stays
      if (runif(1) < 0.15) {
        insert_pos <- which(base_activities == "Treatment_Plan")
        base_activities <- append(base_activities, c("Extended_Observation", "Specialist_Consult"), insert_pos)
      }
    }
    
    # Generate realistic timestamps
    start_time <- as.POSIXct("2024-01-01") + days(sample(0:365, 1)) + hours(sample(0:23, 1))
    
    # Create time intervals that reflect clinical reality
    if (is_sepsis) {
      # Sepsis patients have more urgent, frequent activities
      base_intervals <- c(0, 0.25, 0.5, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 12, 16, 20, 24, 28, 36, 48, 72, 96, 120, 144)
      intervals <- base_intervals[1:min(length(base_activities), length(base_intervals))]
      if (length(intervals) < length(base_activities)) {
        additional <- seq(from = max(intervals), 
                         length.out = length(base_activities) - length(intervals) + 1,
                         by = 12)[-1]
        intervals <- c(intervals, additional)
      }
    } else {
      # Normal patients have more spaced activities
      intervals <- cumsum(rexp(length(base_activities), rate = 1/4))
    }
    
    timestamps <- start_time + hours(intervals[1:length(base_activities)])
    
    # Assign realistic resources
    resources <- character(length(base_activities))
    for (j in seq_along(base_activities)) {
      activity <- base_activities[j]
      resources[j] <- case_when(
        str_detect(activity, "ED_|Triage") ~ "Emergency_Department",
        str_detect(activity, "Lab_|Blood_") ~ "Laboratory",
        str_detect(activity, "ICU|Vasopressor|Mechanical") ~ "ICU",
        str_detect(activity, "Vitals|Assessment") ~ "Nursing",
        str_detect(activity, "Physician|Specialist") ~ "Physician",
        str_detect(activity, "Discharge") ~ "Discharge_Planning",
        TRUE ~ sample(c("Nursing", "Physician", "Support_Staff"), 1)
      )
    }
    
    # Create patient event log
    patient_events <- tibble(
      case = patient_id,
      activity = base_activities,
      timestamp = timestamps,
      resource = resources,
      sepsis_label = as.integer(is_sepsis),
      patient_age = sample(18:95, 1),
      gender = sample(c("M", "F"), 1),
      comorbidities = sample(0:3, 1)
    )
    
    all_events <- append(all_events, list(patient_events))
  }
  
  # Combine all patients
  combined_data <- bind_rows(all_events) %>%
    arrange(case, timestamp) %>%
    mutate(
      lifecycle = "complete",
      activity_instance_id = row_number()
    )
  
  return(combined_data)
}

# Generate our dataset
sepsis_data <- create_realistic_sepsis_data(n_patients = 80, sepsis_rate = 0.3)

# Display basic information
cat("📊 Dataset Overview:\n")
cat(sprintf("  • Patients: %d\n", n_distinct(sepsis_data$case)))
cat(sprintf("  • Events: %d\n", nrow(sepsis_data)))
cat(sprintf("  • Sepsis cases: %d (%.1f%%)\n", 
            sum(sepsis_data$sepsis_label), 
            mean(sepsis_data$sepsis_label) * 100))
cat(sprintf("  • Activities: %d unique\n", n_distinct(sepsis_data$activity)))
```

## Step 1: Data Loading and Validation

```{r}
# Initialize data loader
loader <- EventLogLoader$new()

# Load and validate the data
event_log <- loader$load_from_data_frame(sepsis_data)

# Validate data structure
is_valid <- loader$validate_event_log(event_log)
cat(sprintf("✅ Data validation: %s\n", ifelse(is_valid, "PASSED", "FAILED")))

# Get comprehensive statistics
stats <- loader$get_basic_statistics(event_log)
print(stats)
```

### Data Quality Assessment

```{r}
# Analyze data quality
quality_metrics <- event_log %>%
  summarise(
    completeness = mean(!is.na(activity)) * 100,
    timestamp_coverage = mean(!is.na(timestamp)) * 100,
    resource_coverage = mean(!is.na(resource)) * 100,
    avg_events_per_case = n() / n_distinct(case),
    date_range_days = as.numeric(difftime(max(timestamp), min(timestamp), units = "days"))
  )

cat("📋 Data Quality Metrics:\n")
cat(sprintf("  • Activity completeness: %.1f%%\n", quality_metrics$completeness))
cat(sprintf("  • Timestamp coverage: %.1f%%\n", quality_metrics$timestamp_coverage))
cat(sprintf("  • Resource coverage: %.1f%%\n", quality_metrics$resource_coverage))
cat(sprintf("  • Avg events per patient: %.1f\n", quality_metrics$avg_events_per_case))
cat(sprintf("  • Data spans: %.0f days\n", quality_metrics$date_range_days))
```

## Step 2: Process Mining Analysis

```{r}
# Initialize process miner
miner <- ProcessMiner$new()

# Calculate comprehensive performance metrics
if (requireNamespace("bupaR", quietly = TRUE)) {
  # Convert to bupaR format
  bupar_log <- loader$convert_to_bupar(event_log)
  
  # Discover process patterns
  metrics <- miner$calculate_performance_metrics(bupar_log)
  cat("🔍 Process Mining Results:\n")
  cat(sprintf("  • Cases: %d\n", metrics$n_cases))
  cat(sprintf("  • Events: %d\n", metrics$n_events))
  cat(sprintf("  • Activities: %d\n", metrics$n_activities))
  cat(sprintf("  • Avg case duration: %.1f hours\n", metrics$avg_case_duration))
  cat(sprintf("  • Process variants: %d\n", metrics$n_variants))
  
  # Analyze bottlenecks
  bottlenecks <- miner$identify_bottlenecks(bupar_log, threshold_percentile = 75)
  cat("\n🚨 Process Bottlenecks:\n")
  print(head(bottlenecks, 5))
  
  # Process variants analysis
  variants <- miner$analyze_process_variants(bupar_log, top_n = 5)
  cat("\n🔄 Top Process Variants:\n")
  print(variants)
  
} else {
  cat("⚠️ bupaR not available. Install with: install.packages('bupaR')\n")
  
  # Fallback analysis without bupaR
  metrics <- list(
    n_cases = n_distinct(event_log$case),
    n_events = nrow(event_log),
    n_activities = n_distinct(event_log$activity)
  )
}
```

### Sepsis vs Non-Sepsis Comparison

```{r}
# Compare sepsis and non-sepsis pathways
pathway_comparison <- event_log %>%
  group_by(case, sepsis_label) %>%
  summarise(
    pathway = paste(activity, collapse = " → "),
    duration_hours = as.numeric(difftime(max(timestamp), min(timestamp), units = "hours")),
    n_activities = n(),
    resource_types = n_distinct(resource),
    .groups = "drop"
  )

comparison_stats <- pathway_comparison %>%
  group_by(sepsis_label) %>%
  summarise(
    n_patients = n(),
    avg_duration = mean(duration_hours, na.rm = TRUE),
    median_duration = median(duration_hours, na.rm = TRUE),
    avg_activities = mean(n_activities),
    avg_resources = mean(resource_types),
    .groups = "drop"
  ) %>%
  mutate(
    sepsis_status = ifelse(sepsis_label == 1, "Sepsis", "Non-Sepsis")
  )

cat("📊 Sepsis vs Non-Sepsis Comparison:\n")
print(comparison_stats %>% select(-sepsis_label))
```

## Step 3: Advanced Analytics

```{r}
# Initialize advanced analyzer
advanced <- AdvancedProcessAnalyzer$new(event_log)

# Patient pathway clustering
cat("🔬 Patient Pathway Clustering:\n")
clustering_results <- advanced$cluster_patient_pathways(n_clusters = 4, method = "kmeans")

cat(sprintf("  • Identified %d distinct patient clusters\n", 
           length(unique(clustering_results$clusters$cluster))))

# Display cluster characteristics
cluster_summary <- clustering_results$clusters %>%
  left_join(pathway_comparison %>% select(case, sepsis_label, duration_hours, n_activities), 
           by = "case") %>%
  group_by(cluster) %>%
  summarise(
    n_patients = n(),
    sepsis_rate = mean(sepsis_label, na.rm = TRUE),
    avg_duration = mean(duration_hours, na.rm = TRUE),
    avg_activities = mean(n_activities, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n📈 Cluster Characteristics:\n")
print(cluster_summary)
```

### Clinical KPIs

```{r}
# Calculate clinical key performance indicators
kpis <- advanced$calculate_clinical_kpis()
cat("📊 Clinical KPIs:\n")
cat(sprintf("  • Average length of stay: %.1f hours\n", kpis$avg_los))
cat(sprintf("  • Case completion rate: %.1f%%\n", kpis$completion_rate * 100))
cat(sprintf("  • Activity efficiency: %.1f%%\n", kpis$activity_efficiency * 100))
cat(sprintf("  • Resource utilization: %.1f%%\n", kpis$resource_utilization * 100))
```

### Predictive Risk Modeling

```{r}
# Predictive monitoring for sepsis risk
predictions <- advanced$predictive_monitoring(
  target_activity = "Sepsis_Confirmed",
  threshold = 0.4
)

risk_summary <- predictions %>%
  summarise(
    total_cases = n(),
    high_risk_cases = sum(prediction == "High Risk", na.rm = TRUE),
    avg_risk_score = mean(risk_score, na.rm = TRUE),
    risk_accuracy = if(sum(!is.na(risk_score)) > 0) {
      # Simple accuracy check (this would need actual outcomes in real data)
      "Available with outcome data"
    } else {
      "N/A"
    }
  )

cat("🎯 Predictive Risk Analysis:\n")
cat(sprintf("  • Total cases analyzed: %d\n", risk_summary$total_cases))
cat(sprintf("  • High-risk predictions: %d\n", risk_summary$high_risk_cases))
cat(sprintf("  • Average risk score: %.3f\n", risk_summary$avg_risk_score))
```

## Step 4: LLM-Powered Clinical Insights

```{r eval=FALSE}
# Note: This section requires an OpenRouter API key
api_key <- Sys.getenv("OPENROUTER_API_KEY")

if (nchar(api_key) > 0) {
  # Initialize LLM analyzer
  llm <- LLMAnalyzer$new(api_key = api_key)
  
  # Prepare comprehensive analysis summary
  analysis_summary <- list(
    patient_metrics = list(
      total_patients = n_distinct(event_log$case),
      sepsis_cases = sum(event_log$sepsis_label > 0),
      sepsis_rate = mean(event_log$sepsis_label)
    ),
    process_metrics = metrics,
    bottlenecks = head(bottlenecks, 3),
    clusters = clustering_results$summary,
    kpis = kpis,
    risk_analysis = risk_summary
  )
  
  # Generate clinical insights
  cat("🤖 Generating AI-powered clinical insights...\n")
  
  clinical_response <- llm$generate_clinical_report(
    process_data = analysis_summary,
    model = "anthropic/claude-3.5-sonnet"
  )
  
  cat("📝 Clinical Insights:\n")
  cat(clinical_response$content)
  
  # Multi-model analysis for consensus
  multi_model_analysis <- llm$analyze_with_multiple_models(
    process_results = analysis_summary,
    models = c("anthropic/claude-3.5-sonnet", "openai/gpt-4o"),
    clinical_context = "sepsis"
  )
  
  cat("\n🎭 Multi-Model Analysis Complete\n")
  cat(sprintf("  • Models consulted: %d\n", length(multi_model_analysis)))
  
} else {
  cat("ℹ️ Set OPENROUTER_API_KEY environment variable to enable LLM analysis\n")
}
```

## Step 5: Comprehensive Reporting

```{r}
# Initialize report generator
generator <- ReportGenerator$new()

# Prepare comprehensive report data
report_data <- list(
  title = "Sepsis Patient Journey Analysis",
  subtitle = "Process Mining and AI-Powered Clinical Insights",
  date = Sys.Date(),
  author = "Clinical Process Mining Team",
  organization = "Healthcare Analytics Department",
  
  executive_summary = sprintf(
    "Analysis of %d sepsis cases reveals distinct progression patterns with %.1f%% sepsis rate. Key findings include identification of %d patient clusters and critical bottlenecks in %s processes.",
    n_distinct(event_log$case),
    mean(event_log$sepsis_label) * 100,
    length(unique(clustering_results$clusters$cluster)),
    nrow(bottlenecks)
  ),
  
  methodology = list(
    data_source = "Electronic Health Records",
    analysis_period = sprintf("%s to %s", min(event_log$timestamp), max(event_log$timestamp)),
    patients_analyzed = n_distinct(event_log$case),
    process_mining_approach = "bupaR-based discovery with advanced analytics",
    ai_models = "Multi-model LLM analysis via OpenRouter"
  ),
  
  key_findings = list(
    process_variants = ifelse(exists("variants"), nrow(variants), "Analysis pending"),
    bottlenecks_identified = nrow(bottlenecks),
    risk_factors = list(
      "Delayed antibiotic administration",
      "Laboratory processing delays", 
      "ICU bed availability",
      "Sepsis protocol adherence"
    ),
    patient_clusters = length(unique(clustering_results$clusters$cluster))
  ),
  
  clinical_metrics = list(
    avg_los = kpis$avg_los,
    completion_rate = kpis$completion_rate,
    efficiency = kpis$activity_efficiency,
    resource_utilization = kpis$resource_utilization
  ),
  
  recommendations = list(
    immediate = c(
      "Implement automated sepsis screening alerts",
      "Reduce laboratory turnaround time",
      "Standardize sepsis response protocols"
    ),
    short_term = c(
      "Develop predictive risk models",
      "Optimize ICU resource allocation",
      "Enhance staff sepsis training"
    ),
    long_term = c(
      "Integrate AI-powered decision support",
      "Establish continuous process monitoring",
      "Create sepsis quality dashboards"
    )
  )
)

# Generate markdown report
markdown_report <- generator$generate_markdown_report(report_data, include_visualizations = FALSE)

# Display report preview
cat("📄 Report Preview (first 500 characters):\n")
cat(paste0(substr(markdown_report, 1, 500), "...\n"))

# In a real scenario, you would save the report
# writeLines(markdown_report, "sepsis_analysis_report.md")
```

## Key Insights and Clinical Implications

### Process Optimization Opportunities

Based on our analysis, several key optimization opportunities emerge:

1. **Early Detection Enhancement**
   - Implement automated vital sign monitoring
   - Create predictive risk algorithms
   - Establish rapid response triggers

2. **Resource Allocation Improvements**  
   - Optimize laboratory processing workflows
   - Improve ICU bed management
   - Enhance staff allocation during high-risk periods

3. **Protocol Standardization**
   - Develop consistent sepsis response protocols
   - Create decision support tools
   - Implement quality monitoring systems

### Clinical Decision Support

The analysis reveals that:

- **Patient clustering** identifies distinct risk profiles requiring different care approaches
- **Bottleneck analysis** pinpoints specific process improvements with highest impact  
- **Predictive modeling** enables proactive intervention before sepsis development
- **Resource optimization** can reduce delays in critical care delivery

## Next Steps and Implementation

### Immediate Actions
1. **Validate findings** with clinical experts and domain specialists
2. **Test predictive models** with historical data and outcomes
3. **Pilot process improvements** in selected departments
4. **Establish monitoring systems** for ongoing quality assessment

### Future Enhancements
1. **Real-time integration** with EHR systems
2. **Expanded predictive capabilities** using machine learning
3. **Multi-site validation** across different healthcare systems
4. **Patient outcome correlation** with process improvements

## Conclusion

This tutorial demonstrates the power of combining process mining with AI for healthcare analytics. The HealthProcessAI framework provides:

- **Comprehensive process discovery** revealing hidden patterns in care delivery
- **Advanced analytics** for patient stratification and risk prediction  
- **AI-powered insights** for clinical decision support
- **Actionable recommendations** for quality improvement

The R implementation offers seamless integration with existing healthcare analytics workflows while providing the flexibility to customize analysis for specific clinical contexts.

---

*For more information about HealthProcessAI, visit: https://github.com/ki-smile/HealthProcessAI*

*SMAILE Lab: https://smile.ki.se*